\documentclass[11pt,a4paper]{article}

% ── Packages ──────────────────────────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{parskip}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{float}
\usepackage{multirow}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, backgrounds}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!60!black,
    urlcolor=blue!60!black,
    citecolor=blue!60!black
}

% ── Title ─────────────────────────────────────────────────────────────────────
\title{
    \Large\textbf{Essence: A Personal Embedding Cluster Framework}\\[4pt]
    \Large\textbf{for Depth-Optimized Recommendations}
}
\author{
    Bidit Das\\
    \small Independent Researcher $\mid$ Dallas, TX\\
    \small \href{mailto:biditdas18@gmail.com}{biditdas18@gmail.com}
}
\date{June 17, 2026}

% ── TikZ diagram styles ───────────────────────────────────────────────────────
\tikzset{
  mainbox/.style={
    rectangle, rounded corners=4pt,
    minimum width=8cm, minimum height=1.0cm,
    text centered, text width=7.8cm,
    font=\small\bfseries, inner sep=6pt
  },
  navybox/.style={mainbox, fill=blue!25!black, text=white},
  bluebox/.style={mainbox, fill=blue!55, text=white},
  lightbox/.style={mainbox, fill=blue!10, text=blue!30!black,
                   draw=blue!40, line width=0.8pt},
  greenbox/.style={mainbox, fill=green!35!black, text=white},
  brownbox/.style={mainbox, fill=orange!40!brown, text=white},
  warmbox/.style={mainbox, fill=orange!15, text=orange!40!brown,
                  draw=orange!40!brown, line width=0.8pt},
  purplebox/.style={mainbox, fill=violet!60!black, text=white},
  clusterbox/.style={
    rectangle, rounded corners=3pt,
    minimum width=2.3cm, minimum height=0.9cm,
    text centered, text width=2.1cm,
    font=\small\bfseries,
    fill=green!15, draw=green!40!black, line width=0.8pt,
    text=green!30!black
  },
  arrow/.style={-Stealth, thick, blue!40!black}
}

\begin{document}
\maketitle

% ── Abstract ──────────────────────────────────────────────────────────────────
\begin{abstract}
We present \textbf{Essence}, a personal embedding cluster framework for
recommendation that operates exclusively within a single user's engagement
history, without consulting cross-user interaction data. Consumed items are
embedded using a pre-trained sentence encoder and clustered into $K$ semantic
centroids via $K$-means; recommendations are generated by retrieving items
similar to the centroid closest to the user's most recent activity. We evaluate
Essence on two datasets, Last.fm-1K (music) and Amazon Books (e-commerce),
against three baselines: a non-personalized popularity baseline, a collaborative
filtering baseline (item-based $k$-nearest-neighbors over the user-item
interaction matrix), and a content-based average-embedding baseline. Evaluation
uses a chronological train/test split per user, consistent with the model's use
of recent interaction history, and standard Recall@K. On long-tail items
(those with a single training-set interaction), the non-personalized popularity
baseline achieves exactly zero recall on both datasets, by construction, since
globally popular items are never singletons. Item-based collaborative filtering
recovers long-tail items only on the denser Amazon Books dataset, where it
marginally exceeds the content baseline; on the sparser Last.fm-1K it recovers
none under deterministic ranking, matching the popularity baseline. On both,
its overall recall falls well below the content baseline. Essence achieves the
highest long-tail recall of all systems on both datasets: Long-Tail Recall@10
of $0.0059$ on Last.fm-1K ($2.01\times$ the content baseline) and $0.0197$ on
Amazon Books ($1.59\times$ the content baseline), while also achieving the
highest overall Recall@10 among the personalized methods. On Last.fm-1K, an ablation on active cluster selection shows that replacing the recency-based mechanism with a static mean-embedding query collapses Essence's long-tail advantage, indicating the recency mechanism, not the clustering step alone, is responsible for the effect. We discuss Essence as a candidate-generation method suited to a dedicated
discovery surface rather than a general-purpose ranking replacement.
Code: \url{https://github.com/biditdas18/essence}
\end{abstract}

% ── 1. Introduction ───────────────────────────────────────────────────────────
\section{Introduction}

Recommendation systems built on collaborative filtering (CF) derive their
signal from cross-user interaction patterns: a user's future preferences are
predicted from the behavior of users who interacted similarly in the past.
This approach performs well on standard benchmarks, particularly on metrics
dominated by popular items, but it carries a structural property worth
examining directly: items with sparse interaction histories are, by
construction, underrepresented in any signal derived from cross-user
co-occurrence. A user with a narrow, idiosyncratic taste profile
(engaging consistently with low-interaction items across a domain) receives
recommendations shaped by aggregate similarity to other users, not by the
specific combination of properties their own history reflects.

This work originated from a practical observation in music recommendation:
users with precise, narrow taste profiles are reliably steered toward
broadly popular artists that are only loosely similar, while artists matching
the specific texture of that taste remain unsurfaced regardless of how long
the user searches the catalog manually. This pattern motivated the central
question addressed here: can a recommendation signal be constructed using
only a single user's own history, without any cross-user data, and if so,
does it recover items that collaborative methods structurally cannot?

We propose Essence, a framework that embeds a user's consumed items into a
semantic space using a pre-trained sentence encoder, clusters them into $K$
interest profiles via $K$-means, and generates recommendations by retrieving
items similar to the centroid closest to the user's recent activity. We refer
to this as the \emph{peer-isolation property}: no cross-user interaction graph
is consulted at any stage. This is a deliberate architectural constraint, not
an oversight: it is the mechanism under test. We describe the resulting objective as depth-optimized: the goal is to retrieve items deep in a user's long-tail interest space, rather than to maximize aggregate engagement on broadly popular items.

The contributions of this work are:
\begin{enumerate}[leftmargin=*, itemsep=2pt]
  \item The \textbf{Essence architecture}: a personal recommendation
        framework requiring no collaborative signal and no model training,
        operating entirely on a single user's history.
  \item A \textbf{recency-based active-cluster-selection} mechanism,
        evaluated against a static mean-embedding alternative via ablation
        (Section~\ref{sec:ablation}).
  \item \textbf{Empirical evaluation} on two datasets: Last.fm-1K and
        Amazon Books, under a chronological train/test split, comparing
        Essence against a real collaborative filtering baseline
        (item-based KNN), a non-personalized popularity baseline, and a
        content-based average-embedding baseline.
  \item An empirical finding across two datasets: non-personalized popularity
        achieves exactly zero long-tail recall by construction; item-based
        collaborative filtering recovers long-tail items only on the denser
        dataset (none on the sparser one) and at a large cost to overall recall;
        and Essence achieves the highest long-tail recall of all systems on both.
\end{enumerate}

% ── 2. Related Work ───────────────────────────────────────────────────────────
\section{Related Work}
\label{sec:related}

\subsection{Collaborative Filtering}
\label{sec:cf}
Matrix factorisation methods~\cite{koren2009} and neural
extensions~\cite{he2017} remain dominant in production recommendation. Their
principal limitation is popularity bias: items with sparse interactions are
poorly represented, resulting in near-zero recall for long-tail items. CF is
optimised for the median user: it performs well on what is broadly consumed
and comparatively weakly on what is individually resonant.

\subsection{Multi-Interest Retrieval}
\label{subsec:multi}

MIND~\cite{li2019} introduced dynamic routing to extract multiple interest
capsules from a user's interaction history, explicitly addressing the failure
of a single user vector to represent diverse tastes. ComiRec~\cite{cen2020}
extended this with explicit diversity controls at serving time.

Essence shares the multi-interest intuition with MIND and ComiRec, the
recognition that a user's preferences cannot be collapsed into a single
vector. However, the similarity ends at the surface. The distinction operates
at two levels.

\textbf{Data level.} MIND and ComiRec require population-level interaction
graphs. Their dynamic routing is trained on millions of (user, item,
interaction) triples across the full user base. Cross-user behavioural data
is not optional: it is the mechanism by which interest capsules are learned.
Essence requires no data from any other user, ever. The $K$-means centroids
are derived entirely from one user's own history.

\textbf{Representation level.} Because MIND's item embeddings are trained on
co-occurrence patterns across users, what an item \emph{means} in that space
is partly defined by who else liked it. An obscure track heard by very few
users is underrepresented or invisible in that embedding space, as its
co-occurrence signal is too sparse to be learned reliably. Essence uses a
pre-trained sentence encoder on item metadata only. The embedding captures
what an item \emph{is}, its latent stylistic properties, independent of
how many people have interacted with it. A singleton item and a chart-topping
item are equally representable.

This distinction suggests a hypothesis: Essence should recover long-tail items at higher rates than systems whose embeddings are shaped by co-occurrence sparsity. We do not test this directly against MIND or ComiRec in this work; that comparison is left to future work (Section~\ref{sec:limitations}).

Table~\ref{tab:comparison} summarises the architectural distinctions between
Essence, MIND, ComiRec, and standard CF.

\begin{table}[H]
\centering
\caption{Architectural comparison: Essence vs.\ prior systems.
         The primary distinction is not the clustering mechanism,
         it is the data requirement and what each system is optimised to find.
         The MIND and ComiRec entries reflect architectural expectations from
         their reliance on cross-user co-occurrence; these two methods are not
         empirically evaluated in this work (Section~\ref{sec:results} evaluates
         CF, Content, and Essence).}
\label{tab:comparison}
\begin{tabular}{lcccc}
\toprule
\textbf{Property} & \textbf{CF} & \textbf{MIND} & \textbf{ComiRec}
                  & \textbf{Essence} \\
\midrule
Cross-user data required    & Yes  & Yes  & Yes  & \textbf{No}  \\
Multiple interest profiles  & No   & Yes  & Yes  & \textbf{Yes} \\
Model training required     & Yes  & Yes  & Yes  & \textbf{No}  \\
Long-tail item recovery     & Poor & Poor & Poor & \textbf{Yes} \\
No cross-user data required & No   & No   & No   & \textbf{Yes} \\
Optimised for popular items & Yes  & Yes  & Yes  & \textbf{No}  \\
Optimised for niche items   & No   & No   & No   & \textbf{Yes} \\
Embedding source            & Co-occurrence & Co-occurrence
                            & Co-occurrence & \textbf{Item metadata} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Session-Based and Content-Based Methods}
BERT4Rec~\cite{sun2019} and GRU4Rec~\cite{hidasi2016} model sequential user
behaviour using transformer and recurrent architectures, capturing short-term
context effectively. Content-based filtering relies on item feature similarity
without interaction logs but typically averages preferences into a single
vector, losing intra-user diversity.

Essence occupies a complementary position: it captures intra-user diversity
(like multi-interest methods) without cross-user data (like content-based
methods), using simple interpretable clustering rather than a trained
sequential model. We acknowledge that BERT4Rec offers richer sequential
modelling in high-interaction settings; Essence targets privacy-first,
low-infrastructure deployment contexts where such models are not viable.

% ── 3. The Essence Framework ──────────────────────────────────────────────────
\section{The Essence Framework}
\label{sec:framework}

\subsection{Notation}

Table~\ref{tab:notation} defines all mathematical notation used in this
section. For each symbol a plain-English equivalent is provided.

\begin{table}[H]
\centering
\caption{Notation Legend}
\label{tab:notation}
\begin{tabular}{ll}
\toprule
\textbf{Symbol} & \textbf{Plain-English Meaning} \\
\midrule
$H_u$                     & All items user $u$ has consumed (full history) \\
$i_1, i_2, \ldots$        & Individual items in the user's history, in order \\
$\phi(i)$                 & Embedding of item $i$, a 384-dim vector describing its content \\
$E_u$                     & All embeddings stacked into a matrix, one row per item \\
$K$                       & Number of interest clusters (we use $K=3$) \\
$c_1, c_2, c_3$           & The $K$ cluster centres, one per interest profile \\
$c_k^*$                   & Active cluster centre, closest to the user's recent activity \\
$\mu(\cdot)$              & Mean (average) of a set of vectors \\
$\cos(\mathbf{a},\mathbf{b})$ & Cosine similarity: how alike two vectors are \\
$I \setminus H_u$         & All items the user has \emph{not} yet consumed, the candidate pool \\
$R_u$                     & Final ranked recommendation list for user $u$ \\
$M$                       & Number of recommendations to return (we use $M=10$) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Problem Formulation}
Given a user's history $H_u$ (all items they have consumed), the
goal is to produce a ranked list $R_u \subseteq I \setminus H_u$ of items
they have not yet consumed but are likely to enjoy. Essence accomplishes this
using only $H_u$ and a shared embedding function $\phi: I \to \mathbb{R}^d$.
No data from any other user is used at any stage.

\subsection{Item Embedding}
Each item is described by a short text string encoding its content properties.
For music: track name and artist name. For books: title and author name. We
convert this text into a 384-dimensional vector using the pre-trained sentence
encoder \texttt{all-MiniLM-L6-v2}~\cite{wang2020}. Items that are
semantically similar produce similar vectors; items that are stylistically
distant produce different vectors. Critically, this embedding is derived from
item content, not from interaction patterns. A niche item and a popular item
are equally representable. All embeddings are computed once offline and cached.

Formally, $\phi(i)$ is the embedding of item $i$. For user $u$ with history
$H_u = \{i_1, \ldots, i_n\}$ we compute the embedding matrix
$E_u \in \mathbb{R}^{n \times d}$, where each row is the embedding of one
item.

\subsection{Interest Profile Clustering}
We apply $K$-means clustering to $E_u$ with $K=3$, yielding centroids
$\{c_1, c_2, c_3\}$. Each centroid represents a distinct interest profile:
the semantic centre of gravity of one coherent facet of the user's taste. A
user who consumes jazz, indie rock, and ambient electronica might end up with
one centroid per genre. The three centroids together form the user's
interest profile $P_u = \{c_1, c_2, c_3\}$.

This is the core differentiator from average-embedding approaches. The mean
of all embeddings collapses all interests into a single blurred point that
sits between clusters and represents none of them well. $K$-means preserves
the separation between interest modes, enabling targeted retrieval against
each mode independently.

$K=3$ was selected as the minimum number of centroids capable of
representing the tripartite interest structure commonly observed in
consumption behaviour (e.g., mood-based, genre-based, and era-based
clusters). Sensitivity analysis over $K \in \{2, 3, 4, 5\}$ is a
planned extension.

For users with fewer than $K$ items, Essence falls back to the content
baseline (single mean embedding).

\subsection{Active Cluster Selection}
\label{subsec:active}
At inference time, Essence selects the active centroid $c_k^*$ as the
cluster centre closest to the mean embedding of the user's $r$ most recent
training-set interactions, ordered chronologically (default $r=10$):

\begin{equation}
  c_k^* \;=\; \arg\min_k \;\bigl\| c_k - \mu\!\left(\phi(h_{n-r}), \ldots, \phi(h_n)\right) \bigr\|_2
  \label{eq:active}
\end{equation}

This mechanism assumes recent activity is informative about which interest
cluster is currently active. We test this assumption directly via ablation
(Section~\ref{sec:ablation}): replacing the recency-based query with the mean
embedding of the user's entire training history degrades long-tail recall
substantially on Last.fm-1K, indicating that recency carries signal
beyond what the $K$-means clustering step alone provides. The recency
mechanism is retained as the default not as an unexamined design choice but
because the ablation supports it.

\subsection{Recommendation Generation}
Given the active centroid $c_k^*$, Essence retrieves the top-$M$ unseen
items by cosine similarity:

\begin{equation}
  R_u \;=\; \operatorname{top-}M\;\bigl\{\, i \in I \setminus H_u \;:\;
  \cos\!\left(\phi(i),\, c_k^*\right) \bigr\}
  \label{eq:reco}
\end{equation}

This retrieval step operates entirely in embedding space with no collaborative
signal. An item's probability of appearing in $R_u$ is determined solely by
its latent similarity to the user's current interest cluster, not by how
many other users have interacted with it.

\subsection{Architecture Overview}

Figure~\ref{fig:pipeline} illustrates the full Essence pipeline from raw user
history to final recommendations.

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=0.55cm, every node/.style={align=center}]

  \node[navybox]  (hist)   {USER HISTORY $H_u$\\
                             \normalfont\small All items the user has consumed
                             (no cross-user data)};

  \node[bluebox,  below=of hist]  (emb)
        {STEP 1: EMBED EACH ITEM\\
         \normalfont\small \texttt{all-MiniLM-L6-v2} $\cdot$
         input: item content text $\cdot$ output: $\mathbb{R}^{384}$};

  \node[lightbox, below=of emb]   (mat)
        {ITEM EMBEDDING MATRIX $E_u \in \mathbb{R}^{n \times 384}$\\
         \normalfont\small One row per item};

  \node[greenbox, below=of mat]   (km)
        {STEP 2: K-MEANS CLUSTERING $(K=3)$\\
         \normalfont\small Groups items into 3 distinct interest profiles};

  \node[clusterbox, below=of km, xshift=-3.0cm] (c1) {$c_1$\\Profile 1};
  \node[clusterbox, below=of km]                (c2) {$c_2$\\Profile 2};
  \node[clusterbox, below=of km, xshift=+3.0cm] (c3) {$c_3$\\Profile 3};

  \node[brownbox, below=2.0cm of km] (active)
        {STEP 3: ACTIVE CLUSTER SELECTION\\
         \normalfont\small $c_k^* = \arg\min_k \|c_k - \mu(\text{last 10 items})\|_2$};

  \node[warmbox,  below=of active] (ck)
        {ACTIVE CENTROID $c_k^*$\\
         \normalfont\small Current interest mode of the user};

  \node[purplebox, below=of ck]   (ann)
        {STEP 4: COSINE SIMILARITY SEARCH\\
         \normalfont\small Score all $i \in I \setminus H_u$ by
         $\cos(\phi(i), c_k^*)$};

  \node[navybox,  below=of ann]   (out)
        {TOP-$M$ RECOMMENDATIONS $R_u$\\
         \normalfont\small Depth-optimised $\cdot$
         Privacy-preserving $\cdot$ Long-tail aware};

  \draw[arrow] (hist)   -- (emb);
  \draw[arrow] (emb)    -- (mat);
  \draw[arrow] (mat)    -- (km);
  \draw[arrow] (km.south) -- ++(0,-0.3) -| (c1.north);
  \draw[arrow] (km.south) -- ++(0,-0.3) -| (c2.north);
  \draw[arrow] (km.south) -- ++(0,-0.3) -| (c3.north);
  \draw[arrow] (c1.south) |- (active.west);
  \draw[arrow] (c2.south) -- (active.north);
  \draw[arrow] (c3.south) |- (active.east);
  \draw[arrow] (active) -- (ck);
  \draw[arrow] (ck)     -- (ann);
  \draw[arrow] (ann)    -- (out);

\end{tikzpicture}
\caption{Essence recommendation pipeline. No cross-user data is consulted at
any stage (peer-isolation property).}
\label{fig:pipeline}
\end{figure}

% ── 4. Experimental Setup ─────────────────────────────────────────────────────
\section{Experimental Setup}

\subsection{Datasets}

\textbf{Last.fm-1K (music).} Contains listening histories from 992 users
across approximately 19 million events~\cite{celma2010}. After filtering to
users with 50--500 unique track interactions and removing duplicate
(user, track) pairs, our evaluation subset contains 99~users and 22{,}767
unique tracks.

\textbf{Amazon Books (e-commerce).} The Amazon Reviews 2023 Books 5-core
dataset~\cite{hou2024}, which guarantees every user and item has at least
5 global interactions. After filtering to users with 20--200 unique item
interactions and subsampling to 2{,}000 users (seed~42), our evaluation
subset contains 2{,}000 users and 61{,}727 unique items. Item embeddings use
title and author name, following the same metadata-only approach as Last.fm-1K.

\subsection{Train/Test Split}
For each user on both datasets, interactions are sorted chronologically and
split per-user: the first 80\% form the training set, and the final 20\%
form the test set. This chronological split is required for internal
consistency with the active-cluster-selection mechanism
(Section~\ref{subsec:active}), which defines ``recent'' interactions in terms
of training-set recency; a random split would make this definition
incoherent, since a randomly held-out test item could chronologically
precede items retained in training.

Item embeddings are computed once, offline, over the full item catalog
(train $\cup$ test), so that no test item is excluded from the retrieval
candidate pool due to missing embedding coverage. This is independent of the
train/test split decision and applies identically under either splitting
strategy.

\subsection{Long-Tail Definition}
Long-tail items are those with insufficient interaction signal for
collaborative filtering to represent reliably. We define long-tail
operationally per dataset, anchored to each dataset's interaction density:

\begin{itemize}[leftmargin=*, itemsep=3pt]
  \item \textbf{Last.fm-1K.} Items with a training-set interaction count of
        exactly~1 (\emph{singleton tracks}): $16{,}761$ of the $18{,}679$
        items appearing in the training set ($89.7\%$). The full candidate
        pool (train~$\cup$~test) is $22{,}767$ tracks.
  \item \textbf{Amazon Books.} Items with a training-set interaction count of
        exactly~1 (\emph{singleton items}): $42{,}878$ of the $50{,}924$
        items appearing in the training set ($84.2\%$). The full candidate
        pool (train~$\cup$~test) is $61{,}727$ items. Despite the 5-core
        guarantee of at least~5 global interactions, subsampling to
        2{,}000 users reduces most items to a single observed interaction
        in the subset's training split.
\end{itemize}

\textbf{Generalisation principle.} The long-tail threshold should be set at
the point where collaborative filtering loses reliable signal, empirically
where item interaction count falls below the minimum required for stable
co-occurrence estimation. In both datasets evaluated here, the singleton
threshold (train interaction count~$= 1$) proved the most meaningful
boundary: alternative thresholds either collapsed toward the full catalog
(losing discriminative power) or excluded too few items. The key invariant
is that long-tail items are defined on the \emph{training set only}, after
the train/test split, ensuring no leakage from held-out interactions into the
long-tail classification.

LT-Recall@10 is computed only over test items falling in the long-tail set.
Users with no long-tail test items are excluded from this metric.

\subsection{Systems Compared}
\begin{itemize}[leftmargin=*, itemsep=3pt]
  \item \textbf{Random.} Recommends $M$ items drawn uniformly from the
        unseen candidate pool. Establishes the floor against which all other
        systems are compared.
  \item \textbf{Popularity.} Recommends the $M$ globally most-interacted-with
        items the user has not consumed. A non-personalized baseline;
        included for comparison but not treated as collaborative filtering.
  \item \textbf{CF (ItemKNN).} An item-based $k$-nearest-neighbors
        collaborative filtering baseline, computed via cosine similarity over
        the sparse user-item interaction matrix. Recommends items most
        similar, by co-occurrence pattern, to items the user has already
        consumed.
  \item \textbf{Content (Average Embedding).} Computes the mean of all the
        user's item embeddings and retrieves the $M$ most similar unseen
        items by cosine similarity. Standard content-based filtering without
        interest decomposition.
  \item \textbf{Essence ($K=3$).} Clusters user history into $K=3$ centroids,
        selects the active centroid by Equation~\eqref{eq:active}, and
        retrieves top-$M$ unseen items by Equation~\eqref{eq:reco}.
\end{itemize}

\subsection{Metrics}

\textbf{What we are measuring.} Each system produces a ranked list of 10
recommended items per user. We know which items that user actually consumed
in the test set (ground truth). The evaluation asks: how much of that ground
truth did the system recover?

This is a retrieval problem, not a classification problem. All systems are
evaluated against the \emph{full item catalog} with no negative sampling,
following the full-ranking evaluation protocol~\cite{krichene2020}. Absolute
recall values are low by construction under this protocol: the meaningful
signal is relative comparison across systems evaluated under identical
conditions.

\begin{itemize}[leftmargin=*, itemsep=3pt]
  \item \textbf{Recall@10.} Fraction of the user's test items appearing in
        the top-10 recommendations.
        \[
          \text{Recall@10} = \frac{|\,R_u \cap \text{Test}_u\,|}{|\,\text{Test}_u\,|}
        \]
        Example: 20 test items, 2 hits in top-10. Recall@10 $= 0.10$.
        Averaged across all users.

  \item \textbf{Long-Tail Recall@10 (LT-Recall@10).} Recall@10 restricted to
        singleton test items only.
        \[
          \text{LT-Recall@10} =
          \frac{|\,R_u \cap \text{LT-Test}_u\,|}{|\,\text{LT-Test}_u\,|}
        \]
        A long-tail hit means the system recommended a niche item that this
        specific user genuinely consumed, without any popularity signal
        and without any other user's behaviour to guide it. This metric
        directly tests Essence's core hypothesis.

  \item \textbf{Lift over random.} Recall@10 divided by expected random
        recall ($M / |I|$). Normalises for candidate pool size and makes
        cross-dataset comparison interpretable.
\end{itemize}

\subsection{Implementation}
All systems implemented in Python~3.11. Embeddings via
\texttt{sentence-transformers} (\texttt{all-MiniLM-L6-v2}, 384~dims).
$K$-means via \texttt{scikit-learn} (\texttt{n\_init=10},
\texttt{random\_state=42}).
All ranking steps break score ties deterministically (stable sort by item
index), so reported results are exactly reproducible; this matters for sparse
baselines such as CF (ItemKNN), where many candidate items receive identical
scores.
Code available at:
\url{https://github.com/biditdas18/essence}

% ── 5. Results ────────────────────────────────────────────────────────────────
\section{Results}
\label{sec:results}

\subsection{Last.fm-1K}

Table~\ref{tab:lastfm} reports results on 99 users (chronological 80/20
split per user). All systems rank against the full item catalog of $22{,}767$
unique tracks (train~$\cup$~test); $16{,}761$ of these are long-tail
(training-set singletons). Lift is computed against the theoretical random
expectation ($10/22{,}767 = 0.000439$).

\begin{table}[H]
\centering
\caption{Results on Last.fm-1K (99 users, chronological split, $M=10$,
         $|I|=22{,}767$). Lift computed against theoretical random
         expectation. Best non-random results in \textbf{bold}.
         LT/Content ratios are computed from unrounded recall values.}
\label{tab:lastfm}
\begin{tabular}{lcccc}
\toprule
\textbf{System} & \textbf{Recall@10} & \textbf{Lift}
                & \textbf{LT-Recall@10} & \textbf{LT / Content} \\
\midrule
Random                   & 0.0001          & ---                 & 0.0000          & ---          \\
Popularity               & 0.0023          & $5\times$           & 0.0000          & ---          \\
CF (ItemKNN)             & 0.0018          & $4\times$           & 0.0000          & ---          \\
Content (Avg Embedding)  & 0.0038          & $9\times$           & 0.0029          & $1.0\times$  \\
\textbf{Essence ($K=3$)} & \textbf{0.0119} & $\mathbf{27\times}$ & \textbf{0.0059} & $\mathbf{2.01\times}$ \\
\bottomrule
\end{tabular}
\end{table}

The Popularity baseline scores exactly zero long-tail recall, by construction,
since globally popular items are never singletons. CF (ItemKNN) also recovers
no long-tail items on Last.fm (LT-Recall = $0.0000$): it assigns sparse
singletons co-occurrence scores too weak to rank them reliably above the
candidate field. This metric is sensitive to tie-breaking on sparse data:
under non-deterministic ordering CF's long-tail recall fluctuates around
$0.004$, but resolves to zero once ties are broken deterministically
(Section~4.6), confirming the absence of a reliable long-tail signal here.
Both Content ($0.0029$) and Essence ($0.0059$) recover long-tail items that CF
does not, and Essence's LT-Recall is $2.01\times$ Content's, indicating that
interest decomposition via $K$-means contributes measurably beyond averaging a
user's full history into a single vector. On overall Recall@10, Essence
($0.0119$) leads all personalized methods, with CF ($0.0018$) falling below
even Content ($0.0038$); Essence is the only method that leads on both metrics.

\subsection{Amazon Books}

Table~\ref{tab:amazon} reports results on 2{,}000 users (chronological 80/20
split per user, $|I| = 61{,}727$, $42{,}878$ long-tail items). Lift is
computed against theoretical random expectation ($10/61{,}727 = 0.00016$).

\begin{table}[H]
\centering
\caption{Results on Amazon Books (2{,}000 users, chronological split,
         $M=10$, $|I|=61{,}727$). Lift computed against theoretical random
         expectation. Best non-random results in \textbf{bold}.}
\label{tab:amazon}
\begin{tabular}{lcccc}
\toprule
\textbf{System} & \textbf{Recall@10} & \textbf{Lift}
                & \textbf{LT-Recall@10} & \textbf{LT / Content} \\
\midrule
Random                   & 0.0001          & ---                  & 0.0000          & ---          \\
Popularity               & 0.0021          & $13\times$           & 0.0000          & ---          \\
CF (ItemKNN)             & 0.0031          & $19\times$           & 0.0137          & $1.11\times$ \\
Content (Avg Embedding)  & 0.0173          & $107\times$          & 0.0124          & $1.0\times$  \\
\textbf{Essence ($K=3$)} & \textbf{0.0221} & $\mathbf{136\times}$ & \textbf{0.0197} & $\mathbf{1.59\times}$ \\
\bottomrule
\end{tabular}
\end{table}

The pattern holds: Essence outperforms Content on long-tail recall
($1.59\times$), and Popularity scores zero. Unlike on the sparser Last.fm-1K,
where CF recovered no long-tail items, here CF (ItemKNN) surfaces some: its
long-tail recall ($0.0137$) marginally exceeds Content's ($0.0124$, a
$1.11\times$ margin) despite a substantially lower overall Recall@10 ($0.0031$
vs.\ $0.0173$). This suggests that on a denser dataset
(2{,}000 users vs.\ 99), co-occurrence-based similarity retains some ability
to surface low-interaction items, though at a cost to overall ranking
precision, a tradeoff distinct from Essence's, which improves on both
metrics simultaneously relative to Content.

\subsection{Active Cluster Selection Ablation}
\label{sec:ablation}

Section~\ref{subsec:active} motivates recency-based active cluster selection
on the grounds that it captures session-aware context without requiring a
trained sequential model. We test this directly by comparing it against an
alternative: selecting the active centroid using the mean embedding of the
user's entire training history, rather than the most recent $r=10$
interactions.

\begin{table}[H]
\centering
\caption{Active cluster selection ablation, Last.fm-1K (chronological
         split, $M=10$).}
\label{tab:ablation}
\begin{tabular}{lccc}
\toprule
\textbf{Variant} & \textbf{Recall@10} & \textbf{LT-Recall@10} & \textbf{LT / Content} \\
\midrule
Essence (recency, $r=10$)             & 0.0119 & 0.0059 & $2.01\times$ \\
Essence (mean-select, full history)   & 0.0035 & 0.0005 & $0.16\times$ \\
Content (Avg Embedding, reference)    & 0.0038 & 0.0029 & $1.0\times$  \\
\bottomrule
\end{tabular}
\end{table}

Mean-select performs below even the Content baseline on long-tail recall.
This is consistent with the mechanism collapsing toward Content: selecting
the centroid nearest a user's global mean embedding is functionally similar
to querying with that mean directly, which is what Content already does,
while incurring the additional cost of clustering without a corresponding
benefit. Recency-based selection is retained as the default mechanism on
the basis of this result, conducted on Last.fm-1K; we have not verified
whether this finding generalizes to Amazon Books and note this as a scope
limitation rather than an established cross-domain result.

\subsection{Effect of User-Authored Review Embeddings (Amazon Books)}
\label{sec:pass2}

As a secondary experiment, we tested whether building user taste profiles
from review text (first-person language describing a user's reaction to
consumed items) rather than item metadata changes Essence's relative
performance. Review coverage was complete for the 2{,}000-user Amazon Books
subset.

\begin{table}[H]
\centering
\caption{Amazon Books: metadata embeddings (Pass~1) vs.\ user-review
         embeddings (Pass~2). Pass~2 builds taste profiles from the user's
         own review text; candidates are scored using metadata embeddings.}
\label{tab:amazon_pass2}
\begin{tabular}{llccc}
\toprule
\textbf{Pass} & \textbf{System} & \textbf{Recall@10}
              & \textbf{LT-Recall@10} & \textbf{LT / Content} \\
\midrule
\multirow{3}{*}{Pass 1 (metadata)}
  & CF (ItemKNN)    & 0.0031 & 0.0137 & ---          \\
  & Content (Avg)   & 0.0173 & 0.0124 & $1.0\times$  \\
  & Essence ($K=3$) & 0.0221 & 0.0197 & $1.59\times$ \\
\midrule
\multirow{3}{*}{Pass 2 (user review)}
  & CF (ItemKNN)    & 0.0031 & 0.0137 & ---          \\
  & Content (Avg)   & 0.0027 & 0.0012 & $1.0\times$  \\
  & Essence ($K=3$) & 0.0041 & 0.0034 & $2.83\times$ \\
\bottomrule
\end{tabular}
\end{table}

Absolute recall degrades substantially in Pass~2. Two factors likely
contribute. First, an embedding-space mismatch: taste profiles in Pass~2
are built from review-text embeddings, while candidate items are still
scored using metadata embeddings, and comparing vectors across two distinct
embedding spaces is less discriminative than comparing within one. Second,
review text is a noisier input than structured metadata: a product review
may describe shipping conditions or personal circumstances unrelated to the
item's content, and an embedding built from off-topic text does not
faithfully represent either the item or the user's response to it.

Essence's relative advantage over Content persists in Pass~2
($2.83\times$ vs.\ $1.59\times$ in Pass~1, both above parity), suggesting the
$K$-means clustering mechanism extracts usable interest structure even
under degraded input quality. We do not treat this as a strong claim, given
the absolute recall values in Pass~2 are low enough that the comparison
carries limited statistical weight; we report it as a directional
observation and a motivation for the embedding-input-curation discussion in
Section~\ref{sec:limitations}.

\subsection{Cross-Dataset Pattern}

Three findings hold consistently across both datasets and all conditions
tested:
\begin{enumerate}[leftmargin=*, itemsep=2pt]
  \item The non-personalized Popularity baseline achieves exactly zero long-tail
        recall in every configuration, by construction. Item-based CF (ItemKNN)
        recovers long-tail items only on the denser Amazon Books dataset
        ($0.0137$, marginally above the content baseline at $1.11\times$); on
        the sparser Last.fm-1K it recovers none under deterministic ranking,
        matching Popularity. On both its overall Recall@10 is well below the
        content baseline, and it trails Essence on long-tail recall throughout.
        CF's long-tail recovery is therefore density-dependent, whereas Essence
        leads on long-tail recall on both datasets regardless of density.
  \item Essence outperforms the content-based average-embedding baseline on
        long-tail recall in every primary configuration
        (Sections 5.1--5.2), and the
        ablation in Section~\ref{sec:ablation} indicates this advantage
        depends specifically on recency-based active cluster selection
        rather than on $K$-means clustering alone.
  \item Absolute recall values are low across all systems under
        full-catalog evaluation without negative sampling, a known property
        of this evaluation protocol~\cite{krichene2020} rather than evidence
        of system failure; the relative ordering between systems, evaluated
        under identical conditions, is the meaningful signal.
\end{enumerate}

% ── 6. Discussion ────────────────────────────────────────────────────────────
\section{Discussion}

\subsection{Mechanism: Why Popularity-Driven Discovery Self-Reinforces}

Collaborative filtering's reliance on cross-user co-occurrence creates a
feedback property worth stating plainly: items that already have more
interactions generate more training signal, which improves their
representation in the model, which increases their likelihood of being
recommended, which generates further interactions. Items with few
interactions do not benefit from this loop, regardless of how well they
might match any individual user's taste, because the signal needed to learn
that match does not exist in sufficient quantity. This mechanism is clearest for the non-personalized popularity baseline, which
records exactly zero long-tail recall in Section~\ref{sec:results}. Item-based
CF escapes it only on the denser Amazon Books dataset, where co-occurrence
signal suffices to surface some low-interaction items; on the sparser
Last.fm-1K it too records zero long-tail recall, and on both it pays a large
cost to overall recall. Essence's retrieval step does not depend on this
loop, since item representations are derived from content rather than
interaction frequency: an item's likelihood of being recommended is a
function of embedding similarity to the active cluster, independent of how
many other users have interacted with it.

\subsection{Deployment Positioning: A Candidate-Generation Method, Not a
Ranking Replacement}

The absolute recall values reported in Section~\ref{sec:results}
(approximately 0.01--0.02 under full-catalog evaluation) should be
interpreted in light of how such a system would actually be deployed.
Essence is not proposed as a replacement for a primary, CF-driven
recommendation feed; CF baselines remain effective at surfacing broadly
popular content efficiently, which is appropriate for a general homepage or
default feed. Essence is better understood as a specialized retrieval path:
a candidate-generation method suited to a dedicated discovery surface
(for example, a ``more like this'' or deep-catalog-exploration interface)
where the explicit goal is surfacing items dissimilar from what
population-level signals would otherwise recommend. Framing Essence this
way clarifies its evaluation criteria: it is not competing with CF on
aggregate engagement metrics, but on its ability to recover items that
CF-driven surfaces structurally cannot reach.

\subsection{Privacy and Deployment Constraints}

Because Essence consults no cross-user data at any stage, it is deployable
in settings where collaborative filtering is not viable: privacy-regulated
environments, on-device recommendation without server-side interaction logs,
federated settings where data cannot leave the user's device, and cold-start
contexts where a population-level interaction graph does not yet exist. This
is a direct consequence of the architecture (Section~\ref{sec:framework})
rather than an added feature, and follows from the same data-level
distinction discussed in Section~\ref{subsec:multi}.

% ── 7. Limitations and Future Work ────────────────────────────────────────────
\section{Limitations and Future Work}
\label{sec:limitations}

\begin{itemize}[leftmargin=*, itemsep=3pt]
  \item \textbf{Scale.} Last.fm-1K evaluation uses 99~users. While Amazon
        Books replicates on 2{,}000 users, larger-scale evaluation on both
        datasets would strengthen generalisability claims.
  \item \textbf{$K$ sensitivity.} $K=3$ is used throughout. An ablation over
        $K \in \{2, 3, 4, 5\}$ would characterise sensitivity to this
        hyperparameter across both domains.
  \item \textbf{Active cluster selection.} The recency-based heuristic is
        simple by design. A learned selection mechanism could improve
        session-aware accuracy without requiring cross-user training.
  \item \textbf{Filter bubble risk.} Peer-isolation removes the serendipity
        that cross-user signals can provide. Whether repeated centroid queries
        cause semantic narrowing is an open empirical question.
  \item \textbf{Embedding input curation.} Section~\ref{sec:pass2}
        demonstrates that noisy embedding inputs degrade performance. The
        quality of what enters the embedding determines the quality of what
        Essence can find. Automated curation of embedding input text
        (filtering off-topic or irrelevant content before encoding) is a
        direct practical extension.
  \item \textbf{Audio-based embeddings.} Current embeddings derive from item
        metadata text only. For music, incorporating audio-based
        representations (capturing sonic texture, timbre, rhythm, and
        harmonic structure directly from audio signal) would more precisely
        model the latent acoustic fingerprint underlying individual taste.
        This represents the primary planned extension of Essence: replacing
        or augmenting text embeddings with audio embeddings and measuring
        whether long-tail recall improves when the embedding captures what
        the music actually sounds like rather than what it is called.
  \item \textbf{Cross-space embedding alignment.} Pass~2 results suggest that
        aligning review-text and metadata embedding spaces (via contrastive
        training or a learned projection layer) could unlock the full
        signal of user-authored representations while maintaining retrieval
        coherence.
  \item \textbf{Comparison against multi-interest baselines.} The architectural
        comparison in Section~\ref{subsec:multi} identifies MIND and ComiRec as
        the closest prior systems. A direct empirical evaluation on a dataset
        where pre-trained MIND or ComiRec models are available would establish
        whether the long-tail advantage attributed to Essence's metadata-only
        embeddings persists against a well-tuned multi-interest baseline.
\end{itemize}

% ── 8. Conclusion ─────────────────────────────────────────────────────────────
\section{Conclusion}

We presented Essence, a personal embedding cluster framework that generates
recommendations by clustering a user's item history into $K$ semantic
centroids and retrieving items by similarity to the centroid closest to the
user's recent activity, without consulting any cross-user data. Across two
datasets (Last.fm-1K and Amazon Books), Essence outperforms both a
content-based average-embedding baseline and two collaborative filtering
baselines (a popularity baseline and an item-based KNN baseline) on
long-tail item recovery, under a chronological evaluation split and
standard Recall@K. The non-personalized popularity baseline achieves exactly zero long-tail recall
on both datasets, by construction. The item-based CF baseline recovers some
long-tail items only on the denser Amazon Books dataset (marginally above the
content baseline); on the sparser Last.fm-1K it recovers none, matching the
popularity baseline, while on both it incurs a large cost to overall recall and
remains below Essence. CF's long-tail recovery is thus density-dependent,
whereas Essence leads on long-tail recall regardless of density.

An ablation (Section~\ref{sec:ablation}) indicates that Essence's long-tail
advantage depends on its recency-based active cluster selection mechanism
specifically, not on $K$-means clustering alone: replacing recency-based
selection with a static mean-embedding query degrades performance below the
content baseline. This finding has been verified on Last.fm-1K only, and we
have not yet confirmed whether it holds on Amazon Books.

The long-tail definition used here (items with exactly one training-set
interaction) produced internally consistent results when applied
independently to two datasets with different domains, scales, and
interaction densities, without modification to the threshold or its
derivation. We report this as a useful property of the metric for
evaluating long-tail recovery, while noting that confirming generality
beyond these two datasets would require testing on additional domains.

Limitations of the present evaluation, beyond those above, are discussed in
Section~\ref{sec:limitations}, including dataset scale, sensitivity to the
choice of $K$, and the increased recall degradation observed when taste
profiles are built from noisy, user-authored text (Section~\ref{sec:pass2})
rather than structured item metadata.

% ── References ────────────────────────────────────────────────────────────────
\begin{thebibliography}{9}

\bibitem{li2019}
Li, C., Liu, Z., Wu, M., Xu, Y., Zhao, H., Huang, P., Kang, G., Chen, Q.,
Li, W., \& Lee, D.\ L.\ (2019).
\textit{Multi-Interest Network with Dynamic Routing for Recommendation
at Tmall}.
CIKM 2019.

\bibitem{cen2020}
Cen, Y., et al.\ (2020).
\textit{ComiRec: Controllable Multi-Interest Framework for Recommendation}.
KDD 2020.

\bibitem{sun2019}
Sun, F., et al.\ (2019).
\textit{BERT4Rec: Sequential Recommendation with Bidirectional Encoder
Representations from Transformer}.
CIKM 2019.

\bibitem{he2017}
He, X., et al.\ (2017).
\textit{Neural Collaborative Filtering}.
WWW 2017.

\bibitem{koren2009}
Koren, Y., Bell, R., \& Volinsky, C.\ (2009).
\textit{Matrix Factorization Techniques for Recommender Systems}.
IEEE Computer.

\bibitem{hidasi2016}
Hidasi, B., et al.\ (2016).
\textit{Session-Based Recommendations with Recurrent Neural Networks}.
ICLR 2016.

\bibitem{wang2020}
Wang, W., et al.\ (2020).
\textit{MiniLM: Deep Self-Attention Distillation for Task-Agnostic
Compression of Pre-Trained Transformers}.
NeurIPS 2020.

\bibitem{celma2010}
Celma, O.\ (2010).
\textit{Music Recommendation and Discovery}.
Springer.

\bibitem{hou2024}
Hou, Y., et al.\ (2024).
\textit{Bridging Language and Items for Retrieval and Recommendation}.
arXiv:2403.03952.

\bibitem{krichene2020}
Krichene, W., \& Rendle, S.\ (2020).
\textit{On Sampled Metrics for Item Recommendation}.
KDD 2020.

\end{thebibliography}

\end{document}
